---
title: "Data Wrangling and Visualisation"
author: "DataTeka: Tatjana Kecojevic"
date: "13/02/2018"
logo : DThexagon.png
biglogo : DThexagon.png
#output: slidy_presentation
#output: html_document
#output: beamer_presentation
output: ioslides_presentation
#meta_img: /images/image.jpg
---

```{r, echo = FALSE, warnings=FALSE, message=FALSE}
r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)
```

## Before there was R, there was S!

**R** is a dialect of **S** language that was developed in 1976 by *Rick Becker* and *John Chambers* at the  Bell Laboratories.

*Rick Becker* gave an excellent keynote talk *"Forty Years of S"* at UseR!2016 conference:   
[Rick Becker @ UseR!2016](http://blog.revolutionanalytics.com/2016/07/rick-becker-s-talk.html) where he talked about development of **S** language that gives explanations for many characteristics of **R** as we know it, including "<-" assignment operator.

1993 Bell Labs gave StatSci (later Insightful Corp.) an exclusive license to develop and sell the **S** language. Insightful sold its implementation of the **S** language under the product name **S-PLUS**.

You can read more about the [history of S, R, and S-PLUS](https://www.whoishostingthis.com/resources/s-plus/)

## then, R was born
In early nineties at the University of Aucklandn in the Department of Statistics **R** was created by *Ross Ihaka* and *Robert Gentleman*.

They used [GNU General Public License](http://www.gnu.org/licenses/gpl-2.0.html) to make R open source free software. 

**R**oss Ihaka and **R**obert Gentleman. *R: A language for data analysis and graphics.* Journal of Computational and Graphical Statistics, 5(3):299–314, 1996

Currently **R** is developed by the **R Development Core Team**, of which *John Chambers* is a member.

## Write R Code

To start using **R** you need to:

1) Install [R](https://cran.r-project.org/) [(and RStudio)](https://www.rstudio.com/products/rstudio/download/#download)

2) Launch it and set your working directory: letting R know where to find all of your files.

3) Start writing **R** code!
 
**Tip**:
When start working on a new R code/R Project in [RStudio IDE](https://support.rstudio.com/hc/en-us/sections/200107586-Using-the-RStudio-IDE) use ```File -> New Project```. This way your working directory would be set up when you start a new project and it will save all your files in it. Next time you open your project it would set project's directory as a working directory... It would help you with so much [more](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects).   


## Before Tidyverse R, there is Base R!
When you download and install **R** for the first time, you are installing **the Base R** software. **Base R** contains most of the functions you’ll use on a daily basis: ```mean()```, ```subset()```...

To learn about **R**'s basic operations, data structures and base functions you could look at one of R-Ladies Manchester's handouts: [Introduction to base R](https://tanjakec.github.io/blog/introduction-to-r/).

If you want to access data and code written by other people, you’ll need to install it as a **package**. An **R package** is a bundle of functions (code), data, documentation, vignettes (examples), stored in one neat place.

"In **R**, the fundamental unit of shareable code is the package." [Hadley Wickham](http://r-pkgs.had.co.nz/intro.html)  

## The verse!

An opinionated collection of **R packages** for data science.

[```install.packages("tidyverse")```](https://www.tidyverse.org/)

[```library(tidyverse)```](https://www.tidyverse.org/packages/)

- Have you tried learning data science by reading books? 

[**R for Data Science**](http://r4ds.had.co.nz/) by Garrett Grolemund & Hadley Wickham

- Have you tried learning data science by posting your questions and discussing it with other people within the R community?

[RStudio Community](https://community.rstudio.com/)


## The ```dplyr``` Package: 
provides a “grammar” (the verbs) for data manipulation and for operating on data frames. The **key opertor and the esential verbs** are :

- ```%>%```: **the “pipe” operator** used to connect multiple verb actions together into a pipeline.
- ```select()```: return a subset of the columns of a data frame.
- ```mutate()```: add new variables/columns or transform existing variables.
- ```filter()```: extract a subset of rows from a data frame based on logical conditions.
- ```arrange()```: reorder rows of a data frame according to single or multiple variables.
- ```summarise()``` / ```summarize()```: reduces each group to a single row by calculating aggregate measures. 


## Chicago Data

**Description:** Chicago daily air pollution and death rate data.
A data frame with 7 columns and 5114 rows. Each row refers to one day. The columns are:

• ```death``` total deaths (per day).

• ```pm10median``` median particles in 2.5-10 per cubic m

• ```pm25median``` medianparticles < 2.5 mg per cubic m (more dangerous).

• ```o3median``` Ozone in parts per billion

• ```so2median``` Median Sulpher dioxide measurement

• ```time``` time in days

• ```tmpd``` temperature in fahrenheit


## 1st look at the data: ```dim()``` & ```head()```

```{r, warnings=FALSE, message=FALSE}
# install.packages("gamair")
library(gamair)
data(chicago)
dim(chicago)
head(chicago)
```

## Examine the structure of the data: ```str()```
```{r}
str(chicago) 
```
The output could look messy  and it might not fit the screen if you're dealing with a big data set that has lots of variables!

## Do it in a tidy way: ```glimpse()```

```{r, warnings=FALSE, message=FALSE}
suppressPackageStartupMessages(library(dplyr))
glimpse(chicago) 
```

## The pipeline operater: ```%>%```

**Left Hand Side (LHS)**\ \ \  ```%>%``` \ \ \ **Right Hand Side (RHS)**

```x %>% f(..., y)``` \
\ \ \ \ \ \ \ \ ```f(x,y)```

The "pipe" passes the **result** of the **LHS** as the 1st operator argument of the **function** on the **RHS**

```3 %>% sum(4)``` \ \ \ \ <==>\ \ \ \  ``` sum(3, 4)```

```%>%```\ \  is very practical for chaining together multiple ```dplyr``` functions in a sequence of operations.

##```select()``` 

- ```starts_with("X")``` every name that starts with "X".
- ```ends_with("X")``` every name that ends with "X".
- ```contains("X")``` every name that contains "X".
- ```matches("X")``` every name that matches "X", where "X" can be a regular expression.
- ```num_range("x", 1:5)```  the variables named x01, x02, x03, x04, x05.
- ```one_of(x)``` => every name that appears in x, which should be a character vector.
```{r, echo = FALSE}
head(chicago, n = 2) 
```

##Select your variables

```{r, echo = TRUE}
chicago_air_measurements <- select(chicago, ends_with("median"))
head(chicago_air_measurements, n = 1)
chicago_air_pm <- chicago[c("pm10median", "pm25median")]
head(chicago_air_pm, n = 1)
chicago_air_pm2 <- select(chicago, starts_with("pm"))
head(chicago_air_pm2, n = 1)
```

##```mutate()```
For example, it would allow you to add to the data frame ```df``` a new column, ```z```, which is the multiplication of the columns ```x``` and ```y```: 

\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ ```mutate(df, z = x * y)```

Let us convert °F into °C:\ \ \ \  T(°C) = (T(°F) - 32) × 5/9

```{r, echo = TRUE}
chicago2 <- mutate(chicago, tmpdc = round((tmpd - 32) / 1.8, digits = 1)) 
head(chicago2, n = 3)
```

##```filter()```
There is a set of logical operators in **R** that you can use inside ```filter()```:

- ```x < y```: TRUE if ```x``` is less than ```y```
- ```x <= y```: TRUE if ```x``` is less than or equal to ```y```
- ```x == y```: TRUE if ```x``` equals ```y```
- ```x != y```: TRUE if ```x``` does not equal ```y```
- ```x >= y```: TRUE if ```x``` is greater than or equal to ```y```
- ```x > y```: TRUE if ```x``` is greater than ```y```
- ```x %in% c(a, b, c)```: TRUE if ```x``` is in the vector ```c(a, b, c)```

##Filter your data

```{r, echo = TRUE}
high_death <- filter(chicago2, death > 200) 
high_death
high_temp_death <- filter(chicago2, death > 200 & tmpdc >= 30)
high_temp_death
```

##```arrange()```
is used to reorder rows of a **d**ata **f**rame (df) according to one of the variables/columns.

- If you pass ```arrange()``` a character variable, **R** will rearrange the rows in alphabetical order according to values of the variable. 
- If you pass a factor variable, **R** will rearrange the rows according to the order of the levels in your factor (running ```levels()``` on the variable reveals this order).


## Arranging your data
```{r, echo = TRUE}
low_2_high <- arrange(chicago, death)
head(low_2_high, n = 4)
high_2_low <- arrange(chicago, desc(death))
head(high_2_low, n = 4)
```

##```summarise()```

- uses the same syntax as ```mutate()```, but the resulting dataset consists of a single row instead of an entire new column in the case of ```mutate()```. 

- builds a new dataset that contains only the summarising statistics.

Let us use ```summarise()``` to print out a summary of chicago data containing two variables: max_detht and the max_tmpd
```{r, echo = TRUE}
summarise(chicago, max_deth = max(death), max_tmpd = max(tmpd))
```

##```%>%``` all up!


```{r, echo = TRUE}
chicago_pipe <- chicago %>%
  filter(!is.na(pm10median) & !is.na(so2median)) %>%
  mutate(tmpdC = round((tmpd - 32) / 1.8, digits = 1))
plot(chicago_pipe$tmpdC, chicago_pipe$death, cex = 0.5, col = "red")
```

## grammer of graphics
Enables you to specify building blocks of a plot and to combine them to create graphical display you want. There are 8 building blocks:

- data
- aesthetic mapping
- geometric object
- statistical transformations
- scales
- coordinate system
- position adjustments
- faceting

## ```ggplot()```
```{r, echo = TRUE}
library(ggplot2)
ggplot(chicago_pipe, aes(x = tmpdC, y = death)) +
  geom_point(col ="red")
```

## adding layers to your ```ggplot()```

```{r, echo = TRUE, eval = FALSE}
ggplot(chicago_pipe, aes(x = tmpdC, y = death, col = "red")) +
  geom_point(alpha = 0.2) +
  geom_smooth(col = "blue") +
  labs (title= " death vs temperature ", 
        x = "°C", y = "death") +
  theme(legend.position = "none", 
        panel.border = element_rect(fill = NA, 
                                    colour = "black",
                                    size = .75),
        plot.title=element_text(hjust=0.5))
```


## Voila
```{r, echo = FALSE, warnings=FALSE, message=FALSE}
ggplot(chicago_pipe, aes(x = tmpdC, y = death, col = "red")) +
  geom_point(alpha = 0.2) +
  geom_smooth(col = "blue") +
  labs (title= " death vs temperature ", 
        x = "°C", y = "death") +
  theme(legend.position = "none", 
        panel.border = element_rect(fill = NA, 
                                    colour = "black",
                                    size = .75),
        plot.title=element_text(hjust=0.5))
```

## useful links: 

cheatsheets:

- [data-wrangling-cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

- [ggplot2-cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

websites:

- [tidyverse, visualization, and manipulation basics](https://www.rstudio.com/resources/webinars/tidyverse-visualization-and-manipulation-basics/)

- [ggplot part of tidy verse](http://ggplot2.tidyverse.org/index.html)

- [Introduction to R graphics with ggplot2](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html#introduction)

## Your turn!

1) upload Daily Mortality Weather and Pollution Data for Chicago: **```chicagoNMMAPS```** available from **```dlnm```** package.
2) have a glance at the data.
3) what are the questions you could ask; could you provide the answers to them?

**There is a chalange:**

```dplyr```'s ```group_by()``` function enables you to group your data. It allows you to create a separate df that splits the original df by a variable.

Knowing about ```group_by()``` function, coud you compute the average pollutant level by month and visualise your result?

## Possible Solution: code
```{r, echo = TRUE, eval = FALSE}
# install and open `dlnm' package and access the data
install.packages("dlnm")
library(dlnm)
data("chicagoNMMAPS")

# group data by month and calculate average monthly polution
my_ch <- chicagoNMMAPS %>%
  group_by(month) %>%
  summarise(pm10 = mean(pm10, na.rm = TRUE))

# visualise the information
ggplot(my_ch, aes(x=month, y = pm10)) +
  geom_line() + geom_point(col = "red") +
  xlab("Month") + ylab("average pm10") +
  scale_x_continuous(breaks = seq(1, 12, 1), labels = seq(1, 12, 1))
```

## Avoid Chicago in spring and summer!
```{r, echo = FALSE, warnings=FALSE, message=FALSE}
# install and open `dlnm' package and access the data
# install.packages("dlnm")
library(dlnm)
data("chicagoNMMAPS")

# group data by month and calculate average monthly polution
my_ch <- chicagoNMMAPS %>%
  group_by(month) %>%
  summarise(pm10 = mean(pm10, na.rm = TRUE))

# visualise the information
ggplot(my_ch, aes(x=month, y = pm10)) +
  geom_line() + geom_point(col = "red") +
  xlab("Month") + ylab("average pm10") +
  scale_x_continuous(breaks = seq(1, 12, 1), labels = seq(1, 12, 1))
```